---
title: "Reproducible Research: Peer Assessment 1"
author: "Mark Salib"
date: "8 April 2017"
output: 
  html_document: 
    keep_md: true 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(httr)
library(lubridate)
library(dplyr)
library(magrittr)
library(ggplot2)
library(lubridate)

GET("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip" , write_disk("Factivity.zip", overwrite = TRUE))
unzip("Factivity.zip")

```

###Loading and preprocessing the data
```{r dataload}

dat <- read.csv("activity.csv")

dat$date <- as.Date(dat$date)

dat <- dat %>% 
        mutate( day = as.numeric(day(dat$date))) %>%
        mutate(steps = as.numeric(steps))

```

##What is mean total number of steps taken per day?
  
1- Calculate the total number of steps taken per day
  
```{r totsteps}

dat %>% 
        group_by(date) %>%
        summarise(totSteps = sum(steps))


```
  
  
2- Make a histogram
  
```{r histogram}

histdat<- dat %>% 
          group_by(date) %>%
          summarise(totSteps = sum(steps)) 

hist(histdat$totSteps, main = "Histogram of the total steps taken per day ignoring NA", xlab = "Steps", breaks = 16, col = "wheat")
        


```
  
  
3- calculate mean and median
  
```{r means}


dat %>% 
        group_by(date) %>%
        summarise(meanSteps = mean(steps), medianSteps = median(steps)) 
        


```


## What is the average daily acitivity pattern?

```{r activitygraph}


plotdat <- dat %>% 
                group_by(interval) %>%
                summarise(meanSteps = mean(steps, na.rm = TRUE)) 
        
ggplot(plotdat, aes(interval, meanSteps)) + geom_line() +
        ggtitle("Mean steps per daily interval") +
        xlab("Daily interval") + 
        ylab("Mean steps")

```
  

2- Find busiest interval  
  
```{r busyint}


dat %>% 
        group_by(interval) %>%
        summarise(meanSteps = mean(steps, na.rm = TRUE)) %>%
        filter(meanSteps == max(meanSteps))
        


```

##Imputing missing values
  
 1- Count how many missing values 
  
```{r countna}

dat %>% summarise(naCount = sum(is.na(steps)))
        
```

2- Devise strategy for filling in the missing values

na values will be replaced by average for that 5 minute interval
 
```{r missingna}

datna.rm <- dat %>%
        group_by(interval) %>%
        summarise(meanSteps = mean(steps, na.rm =TRUE))


datNew <- dat %>% mutate(newSteps = ifelse(is.na(steps) , datna.rm$meanSteps[match(interval, datna.rm$interval)] , steps))


histdat<- datNew %>% 
          group_by(date) %>%
          summarise(totSteps = sum(newSteps, na.rm = TRUE)) 

```

Calculate the total, mean and median of the adjusted data
 
```{r newmeanmedian}

datNew %>% 
        group_by(date) %>%
        summarise(totSteps = sum(newSteps), meanSteps = mean(newSteps), medianSteps = median(newSteps) ) 


        
```

Compare histograms


```{r comparehisto}

histdat_before<- dat %>% 
          group_by(date) %>%
          summarise(totSteps = sum(steps)) 



histdat_after<- datNew %>% 
          group_by(date) %>%
          summarise(totSteps = sum(newSteps)) 


hist(histdat_after$totSteps, main = "Total steps each day", xlab = "Steps", breaks = 16, col = "red" )
hist(histdat_before$totSteps, main = "Total steps each day", xlab = "Steps", breaks = 16, col = "blue", add = TRUE)
legend("topright", c("non-adjusted", "adjusted"), col=c("red", "blue"), lwd=3)
    

```



Whats the final effect of adjusting the NAs?

- More total steps but with same mean which is consistant with the method I used to replace NAs

```{r final effect}

dat %>% summarise(totSteps = sum(steps, na.rm = TRUE), meanSteps = mean(steps, na.rm = TRUE) , medianSteps = median(steps, na.rm = TRUE))

datNew %>%  summarise(totNEWSteps = sum(newSteps), meanNEWSteps = mean(newSteps) , medianNEWSteps = median(newSteps))

 
```


## Are there differences in activity patterns between weekdays and weekends?

- Yes more steps on the weekend when youre not stuck at work or school!

```{r weekends}

library(lubridate)

datNew <- datNew %>% mutate(daytype =ifelse(wday(date) %in% c(6,7) , "weekend" , "weekday" ))

plotdat <- datNew %>% 
                group_by(interval, daytype) %>%
                summarise(meanSteps = mean(newSteps)) 
        
ggplot(plotdat, aes(interval, meanSteps, colour = daytype)) + geom_line() +
        ggtitle("Mean steps per daily interval grouped by weekend/weekeday") +
        xlab("Daily interval") + 
        ylab("Mean steps") +
        facet_wrap(~daytype, ncol = 1, nrow = 2)

```
